---
title: "Change of manager effect in football"
author: "Pascal Albisser"
date: "2023-02-17"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load dataset

Source: <https://www.kaggle.com/datasets/davidcariboo/player-scores>

```{r loading-data}
games <- read.csv("./data/kaggle_transfermarkt/games.csv") %>% 
  mutate(date = as.Date(date))
summary(games)
```

### Look at one team

Take Athletic Bilbao for example:

```{r}

games %>% 
  filter(home_club_id == 621 & competition_type == "domestic_league") %>% 
  arrange(date) %>% 
  select(c("competition_type", "date", "season", "round", "home_club_manager_name", "aggregate", "club_home_name", "club_away_name"))

```

So we need to get rid of the home/away distinction somehow to have a constant timeline. 
- double the dataset
- transform into team of interest vs. opponent

```{r}

games_home <- games %>% 
  rename(team_id = home_club_id,
         team_name = club_home_name,
         team_goals = home_club_goals,
         opponent_goals = away_club_goals,
         team_position = home_club_position,
         team_manager = home_club_manager_name,
         ) %>% 
  select(c("game_id", "team_name", "competition_type", "date", "season", "round", "team_id", "team_goals", "opponent_goals", "team_position", "team_manager", "stadium", "attendance", "referee", "url")) %>%
  mutate(where = "Home") %>% 
  arrange(team_id)

games_away <- games %>% 
  rename(team_id = away_club_id,
         team_name = club_away_name,
         team_goals = away_club_goals,
         opponent_goals = home_club_goals,
         team_position = away_club_position,
         team_manager = away_club_manager_name,
         ) %>% 
  select(c("game_id", "team_name", "competition_type", "date", "season", "round", "team_id", "team_goals", "opponent_goals", "team_position", "team_manager", "stadium", "attendance", "referee", "url")) %>%
  mutate(where = "Away") %>% 
  arrange(team_id)

df_games <- rbind(games_home, games_away)

df_games <- df_games %>% 
  arrange(game_id)

head(df_games)

```
Check again pro team, we should now have a complete list of games without missing games in between.


```{r}
df_games %>% 
  filter(team_id == 621) %>% 
  arrange(date) 
```

## Construct variables

Goal margin as first indicator of performance.

```{r}
df_games <- df_games %>% 
  mutate(goal_margin = team_goals - opponent_goals)
```

```{r}
df_games %>% 
  filter(team_id == 621) %>% 
  arrange(date) %>% 
  mutate(team_manager_last_game = lag(team_manager)) %>% 
  select(c("date", "team_manager", "team_manager_last_game"))

# do this for every team
```

## Plot
```{r}
df_games %>% 
  filter(team_id == 621 & season == 2012) %>% 
  ggplot(mapping = aes(x = date, y = goal_margin, group = team_id)) +
    geom_step() +
    labs(title = "Goal Margin of Athletic Bilbao for Season 2012", y = "Scored - received goals")
```


ggplot()
