---
title: "Change of manager effect in football"
author: "Pascal Albisser"
date: "2023-03-06"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
---

```{r, message = FALSE, echo = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(fixest)
library(kableExtra)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = function() {
  if (knitr::is_latex_output())
    "latex" else "pipe"
})
```

# Introduction

Whenever a football team performs badly oftentimes the first measure taken is the sacking of the teams manager. In the swiss football league for example the FC Sion and his president Christian Constantin are known for their rigorous sacking of managers for not delivering the desired results. Recently (03.03.2023) they suspended their manager Fabio Celestini after six games without win as [srf.ch](https://www.srf.ch/sport/fussball/super-league/aus-nach-nur-6-pflichtspielen-sion-entlaesst-celestini-definitiv) and others reported.

I personally feel this may be common but not a very effective measure. Some teams get better afterwards, some not. Others get better without changing their manager. So why not have a look at some data to scrutinize if changing the manager of a bad performing team has an effect on its position in the league?

# Data Source

To examine this question I take a look at a [football dataset](https://www.kaggle.com/datasets/davidcariboo/player-scores) from kaggle.com. It consists of 60'535 observations of football games scraped from the [www.transfermarkt.co.uk](https://www.transfermarkt.co.uk/). The dataset includes many values among them the name of the current manager. This allows us detect changes of the managers which is our treatment in this case.

```{r loading-data}
games <- read.csv("./data/kaggle_transfermarkt/games.csv") %>% 
  mutate(date = as.Date(date))
summary(games)
```
## Preprocessing

Although the data contain all necessary values it is of inconvenient form for my analysis. I want to have one team per row for example. But the dataset has two in each, namely the home and away team per match. This requires some preprocessing of the data.

```{r}

# seperate observations for home...
games_home <- games %>% 
  rename(team_id = home_club_id,
         team_name = club_home_name,
         team_goals = home_club_goals,
         opponent_goals = away_club_goals,
         team_position = home_club_position,
         team_manager = home_club_manager_name
         ) %>% 
  select(c("game_id", "team_name", "competition_type", "date", "competition_id", "season", "round", "team_id", "team_goals", "opponent_goals", "team_position", "team_manager", "stadium", "attendance", "referee", "url")) %>%
  mutate(where = "Home") %>% 
  arrange(team_id)

# ...and away teams
games_away <- games %>% 
  rename(team_id = away_club_id,
         team_name = club_away_name,
         team_goals = away_club_goals,
         opponent_goals = home_club_goals,
         team_position = away_club_position,
         team_manager = away_club_manager_name
         ) %>% 
  select(c("game_id", "team_name", "competition_type", "date", "competition_id", "season", "round", "team_id", "team_goals", "opponent_goals", "team_position", "team_manager", "stadium", "attendance", "referee", "url")) %>%
  mutate(where = "Away") %>% 
  arrange(team_id)

# combine both
df_games <- rbind(games_home, games_away)

df_games <- df_games %>% 
  arrange(game_id)

# Some 1314 observations have missing values for managers - drop these
df_games <- df_games %>% 
  filter(team_manager != "")

str(df_games)

```


## Construct Variables

I add a column indicating manager changes in comparison to the last game played to detect a treatment. 

```{r}
df_games <- df_games %>% 
  arrange(date) %>% 
  group_by(team_id) %>% 
  mutate(team_manager_last_game = lag(team_manager)) %>% 
  mutate(manager_change = team_manager_last_game != team_manager) %>% 
  ungroup()

kable(head(df_games %>% 
  filter(team_manager != team_manager_last_game) %>% 
  select(team_name, team_manager, team_manager_last_game, manager_change)))
```

## Take spanish league as subset

The dataset contains teams of different football leagues. Some teams take part of other competitions as the Champions League or the domestic cup as well. For reasons of simplicity I restrict my analysis on the games and teams of spanish league (La Liga). This provides us with an enclosed system to look at and more importantly with a constant indicator of performance: the rank on the league table.  

```{r}
spanish_league <- df_games %>% 
  filter(competition_id == "ES1") %>% 
  mutate(round_n = as.numeric(sub(" .*", "", round))) %>% 
  group_by(team_id) %>%
  mutate(games_since_last_change = with(rle(manager_change), sequence(lengths)) * !manager_change) %>%
  ungroup() %>% 
  arrange(desc(date)) %>% 
  group_by(team_id) %>% 
  mutate(games_to_change = with(rle(manager_change), sequence(lengths)) * !manager_change) %>%
  ungroup() %>% 
  arrange(team_id, date)
```

## Overview

In order to get a feeling of the dynamics of the data I plot here all games per team with indicators where a manager change happened (red).

```{r}

spanish_league %>% 
  arrange(team_id, date) %>% 
  #select(date, team_id, team_manager, team_manager_last_game, games_since_last_change, games_to_change, round_n) %>% 
  ggplot(mapping = aes(y = factor(team_name), x = date)) +
    geom_point(shape = "|") +
    geom_point(data = . %>% filter(manager_change), shape = 18, colour = "red", size = 3) +
    labs(title = "Games & manager changes in La Liga per team", subtitle = "Seasons 2011 - 2022", x = "", y = "team")
  
```

As every season every team starts from zero it takes some time until its position is more or less stable. Or put differently: until the rank is a solid representation of the teams ability. The next line chart depicts that apart from round 5 there are on average only changes of about 2 ranks. Therefore we exclude the first 5 rounds from our analysis.

```{r}
spanish_league %>% 
  group_by(team_id) %>% 
  mutate(position_delta = abs(lag(team_position) - team_position)) %>% 
  select(team_id, round_n, season, team_position, position_delta) %>% 
  group_by(round_n) %>% 
  summarise(delta = mean(position_delta, na.rm = TRUE)) %>% 
  ggplot(mapping = aes(y = delta, x = round_n)) +
    geom_line() +
    geom_vline(xintercept = 5) +
    labs(title = "Decline in position changes", y = "average position change", x = "round of season")

```

# The Difference-in-Differences (DID) approach

So far we can think of two main explanations for the possible changes to a teams performance: Time passing by and beeing treated or not, that is have its manager replaced or not. Therefore I am going to build a control group consisting of teams which did not replace their coach and a treatment group which did. This control group then allows me to estimate the changes which would have happened to the treatment group without treatment and then estimate the difference to the actual change. This difference is the effect of the treatment.  

## Prepare treatment group

To measure if the treatment has an effect I take the rank 3 games before the treatment, as well as right at the time of treatment and 3 games after the treatment as indicators. As I only want to scrutinize manager changes due to bad results I filter on negative rank changes before the treatment. In the end the group consists of 57 treatments.

```{r, echo }
df_treatments <- spanish_league %>% 
  group_by(team_id) %>% 
  mutate(position_start = lag(team_position, 4L, order_by = team_id)) %>% 
  mutate(position_before = lag(team_position, 1L, order_by = team_id)) %>% 
  mutate(position_after = lead(team_position, 2L, order_by = team_id)) %>% 
  mutate(position_start_rel = 0) %>% 
  mutate(position_before_rel = position_start - position_before) %>% 
  mutate(position_after_rel = position_start - position_after) %>% 
  ungroup() %>% 
  filter(manager_change & round_n > 5) %>% 
  select(game_id, team_id, date, round_n, season, team_name, team_manager, team_position, team_manager_last_game, position_start_rel, position_before_rel, position_after_rel) %>% 
  filter(position_before_rel < 0) %>%  # only treatments with decrease in position before 
  mutate(group_id = row_number())

df_treatments <- df_treatments %>% 
  pivot_longer(
    cols = starts_with("position_"),
    names_to = "t",
    values_to = "position"
  ) %>% 
  mutate(t = factor(t, levels = c("position_start_rel", "position_before_rel", "position_after_rel"), labels = c("-3 games", "at treatment", "+3 games"))) %>% 
  arrange(team_id, date)

message(length(unique(df_treatments$group_id)), " samples")
  
```


```{r, warning = FALSE}
df_treatments %>%   
  ggplot(mapping = aes(x = t, y = position, group = group_id, colour = "treatment")) +
    geom_point(alpha = 0.1) +
    geom_line(alpha = 0.3) +
    geom_smooth(formula = y ~ x, method = "loess", mapping = aes(group = NA, colour = "loess"), linewidth = 1) +
    labs(title = "Change of position in treatment group", x = "", y = "relative position") +
    scale_colour_manual(name="", values=c("red", "grey"))

```

Just looking at the treatment group itself the manager change seems to have at least a small positive effect. But most of the teams could not make up for the loss of ranks before.

## Prepare control group

As for the control group I want to find teams also decreasing in rank but then decide not to sack their manager. Additionally I also make sure time periods per team do not overlap.

```{r}

df_control <- spanish_league %>%
  group_by(team_id) %>% 
  mutate(position_start = lag(team_position, 4L, order_by = team_id)) %>% 
  mutate(position_before = lag(team_position, 1L, order_by = team_id)) %>% 
  mutate(position_after = lead(team_position, 2L, order_by = team_id)) %>% 
  mutate(position_start_rel = 0) %>% 
  mutate(position_before_rel = position_start - position_before) %>% 
  mutate(position_after_rel = position_start - position_after) %>% 
  mutate(same_manager = lag(team_manager, 4L, order_by = team_id) == team_manager & team_manager == lead(team_manager, 2L, order_by = team_id)) %>%  #--> need to check if same manager in period
  ungroup() 

df_control <- df_control %>%
  filter(position_before_rel < 0) %>%
  filter(same_manager) %>%
  mutate(round_distance = abs(round_n - lead(round_n))) %>%
  filter(round_distance > 3) %>%
  filter(round_n > 5) %>%
  select(game_id, team_id, date, round_n, season, team_name, team_position, team_manager, team_manager_last_game, position_start_rel, position_before_rel, position_after_rel) %>%
  mutate(group_id = row_number()) %>%
  pivot_longer(
    cols = starts_with("position_"),
    names_to = "t",
    values_to = "position"
  ) %>%
  mutate(t = factor(t, levels = c("position_start_rel", "position_before_rel", "position_after_rel"), labels = c("-3 games", "at treatment", "+3 games")))

message(length(unique(df_control$group_id)), " samples")

```


```{r, warning = FALSE}

df_control %>%   
  ggplot(mapping = aes(x = t, y = position, group = group_id, colour = "treatment")) +
    geom_point(alpha = 0.1) +
    geom_line(alpha = 0.3) +
    geom_smooth(formula = y ~ x, method = "loess", mapping = aes(group = NA, colour = "loess"), linewidth = 1) +
    labs(title = "Change of position in control group", x = "", y = "relative position") +
    scale_colour_manual(name="", values=c("red", "grey"))

```

The chart indicates that not changing the manager also leads to better results.

## Checking the parallel trend assumption

In order for the DID-approach to be plausible and make sure the change in the treatment group is due to a causal effect it is important to check the parallel trends assumption. I have to show that without treatment both groups would have gone the same way. For the treatment this is a counter factual - I can not actually observe it because it never happened. And therefore I can not mathematically prove it.

But if the prior trends of both groups are somewhat similar it is likely that both groups would have further decreased in ranks. By finding control observations based on similarity to the treatment observations I already can be quite sure prior trends are trending the same way. The following plot confirms this.

```{r, warning = FALSE, message = FALSE}

df_control$treatment <- 0
df_treatments$treatment <- 1

df_did <- rbind(df_control, df_treatments) %>% 
  mutate(treatment = factor(treatment, levels = c(0,1), labels = c('Control', 'Treatment')))

df_did %>% 
  group_by(t, treatment) %>% 
  summarize(avg_position = mean(position)) %>% 
  ungroup() %>% 
    ggplot(mapping = aes(x = t, y = avg_position, color=factor(treatment), group = treatment))+
      geom_line(size=1)+
      geom_vline(xintercept=2.1) +
      guides(color = guide_legend(title = element_blank())) +
      labs(title = "Trend comparison of control and treatment group", x = "Time", y = "Position (Relative)")
```

In addition most of the circumstances stay constant over the whole period for both groups. All games always start from 0-0, the games are held on the same pitches and even the players are mostly constant due to the fact that transfers are only allowed twice a year between seasons. So we have reason to believe that the untreated group was not object of sudden changes around the time of treatment and both groups are similar. 

## Formula

We can express our assumption in the following formula:

\begin{align*}
Y = \beta_{0} + \beta_{1}ManagerChangedGroup + \beta_{2}AfterManagerChange + \\ 
\beta_{3}ManagerChangedGroup x AfterManagerChange + \varepsilon
\end{align*}

The coefficient <math xmlns="http://www.w3.org/1998/Math/MathML">
  <msub>
    <mi>&#x3B2;</mi>
    <mn>3</mn>
  </msub>
</math> of the interaction term <math xmlns='http://www.w3.org/1998/Math/MathML'>
 <mrow>
  <mi>AfterManagerChange</mi>
  <mo>&#8290;</mo>
  <mi>x</mi>
  <mo>&#8290;</mo>
  <mi>ManagerChangedGroup</mi>
 </mrow>
</math> tells us the difference of the <math xmlns='http://www.w3.org/1998/Math/MathML'>
 <mi>ManagerChangedGroup</mi>
</math> effect before and after the treatment - the difference-in-differences we are looking for.

## Average treatment effect

The explanation of DID above already mentioned I can subtract the change in the untreated group from the change in the treated group to get the effect of changing manager on bad performing teams:

```{r}

df_did$post <- as.numeric(df_did$t == "+3 games")

(mean(df_did[df_did$treatment=="Treatment" & df_did$post==1,]$position)-
mean(df_did$position[df_did$treatment=="Control" & df_did$post==1]))-
(mean(df_did$position[df_did$treatment=="Treatment" & df_did$post==0])-
mean(df_did$position[df_did$treatment=="Control" & df_did$post==0]))
```

This indicates that teams changing their manager on average occupy a lower rank by 0.79 after 3 games than teams that do not.

Now I want to see if the effect is significant. But first let me formulate my hypotheses:

- H<sub>0</sub>: the replacement of a manager __has no effect__ on the performance of a football team
- H<sub>A</sub>: the replacement of a manager __has an effect__ on the performance of a football team

To estimate the treatment effect I use the `feols`-command from the `fixest`-package. This also allows me to add a fixed effect for team and therefore control for all things that are constant per team over years. For example I assume that teams like _FC Barcelona_ and _Real Madrid_ constantly have a larger budget than other teams and therefore probably more means to influence the performance of the team.  

```{r}
did_fe_teams <- feols(fml = position ~ post * treatment | team_id, data = df_did)
etable(did_fe_teams)
```
The result of the fixed-effect OLS indicates we reject the null-hypothesis at the 10% significance level, but not at the more common 5% significance level. With some certainty I can say there is a changing-manager-effect on bad performing teams. But not in the way I believe the it is thought to be: the difference is negative, meaning __teams who keep their coaches rank better by roughly 0.8 ranks__ after 3 games.

# Conclusion

I was able to show that changing the manager on a bad perfoming team does not have the desired effect. Even more: Teams keeping their coaches seem to have better results.

## Comparison

My findings are consistent with a study in the early 2000s on 12 seasons of dutch football results. Allard Bruinshoofd and Bas ter Weel concluded: "turns out that when the manager would not have been forced to resign, performance would have improved more rapidly" (<http://dx.doi.org/10.1016/S0377-2217(02)00680-X>).

Sue Bridgewater finds in her study from 2009 that there is a positive short-term effect after changing the manager, but it does not improve performances in the the long term (<http://dx.doi.org/10.13140/2.1.3696.0328>). This comparison is not made here, but when considering 3 games as short term effect there is some contradiction to my findings. 

But I also call into question the findings of the Bridgewater for the lack of a control group. I think as fourfourtwo.com pointed out correctly in [an article](https://www.fourfourtwo.com/features/soccernomics-does-sacking-manager-actually-make-difference) about teams in a bad period: "whether or not the club sacks its manager, or changes its brand of teacakes, its performance will probably “regress to the mean” – or in ordinary language, return to normal. Simply put, from a low point you are always likely to improve." This again supports my outcome to a certain amount and would probably have changed the outcomes of Sue Bridgewaters study.


## Drawbacks & possible improvements

For sure only looking at ranks might be a very rough measure. Teams could change coaches but then have to play the 3 best teams of the league afterwards. They might play better under the influence of the new manager but still loose. But on the other hand it is what counts in the end. The rank decides about winning the league, participation in international competitions or relegation.

Further research could make us of the [TrueSkill-algorithm](https://www.microsoft.com/en-us/research/publication/trueskilltm-a-bayesian-skill-rating-system/?from=https://research.microsoft.com/apps/pubs/default.aspx?id=67956&type=exact) or take a performance indicator relative to the teams average (as the studies above) to better depict the teams performance. And of course, there are more leagues to look at than just the spanish La Liga.

In terms of methods further research could take better care of the fact that the treatment does not happen at the some point in time but at many different points. This is also called the staggered rollout problem, as stated by Nick Huntington-Klein in his book "The Effect"([Doing Multiple Treatment Periods Right!](https://theeffectbook.net/ch-DifferenceinDifference.html#doing-multiple-treatment-periods-right)). Although I have very carefully choosen my treatment and control samples and shown that they are comparable there might still be some issues with teams beeing treated at different points in time. 

