---
title: "Change of manager effect in football"
author: "Pascal Albisser"
date: "2023-02-17"
output: html_document
---

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load dataset

Source: <https://www.kaggle.com/datasets/davidcariboo/player-scores>

```{r loading-data}
games <- read.csv("./data/kaggle_transfermarkt/games.csv") %>% 
  mutate(date = as.Date(date))
str(games)
```

### Look at one team

Take Athletic Bilbao for example:

```{r}

games %>% 
  filter(home_club_id == 621 & competition_type == "domestic_league") %>% 
  arrange(date) %>% 
  select(c("competition_type", "date", "season", "round", "home_club_manager_name", "aggregate", "club_home_name", "club_away_name"))

```

So we need to get rid of the home/away distinction somehow to have a constant timeline. 
- double the dataset
- transform into team of interest vs. opponent

```{r}

games_home <- games %>% 
  rename(team_id = home_club_id,
         team_name = club_home_name,
         team_goals = home_club_goals,
         opponent_goals = away_club_goals,
         team_position = home_club_position,
         team_manager = home_club_manager_name
         ) %>% 
  select(c("game_id", "team_name", "competition_type", "date", "competition_id", "season", "round", "team_id", "team_goals", "opponent_goals", "team_position", "team_manager", "stadium", "attendance", "referee", "url")) %>%
  mutate(where = "Home") %>% 
  arrange(team_id)

games_away <- games %>% 
  rename(team_id = away_club_id,
         team_name = club_away_name,
         team_goals = away_club_goals,
         opponent_goals = home_club_goals,
         team_position = away_club_position,
         team_manager = away_club_manager_name
         ) %>% 
  select(c("game_id", "team_name", "competition_type", "date", "competition_id", "season", "round", "team_id", "team_goals", "opponent_goals", "team_position", "team_manager", "stadium", "attendance", "referee", "url")) %>%
  mutate(where = "Away") %>% 
  arrange(team_id)

df_games <- rbind(games_home, games_away)

df_games <- df_games %>% 
  arrange(game_id)

# Some 1314 observations have missing values for managers - drop these
df_games <- df_games %>% 
  filter(team_manager != "")

head(df_games)

```
Check again pro team, we should now have a complete list of games without missing games in between.

```{r}
df_games %>% 
  filter(team_id == 621) %>% 
  arrange(date) 
```


## Construct variables

Goal margin as first indicator of performance.

```{r}
df_games <- df_games %>% 
  mutate(goal_margin = team_goals - opponent_goals)
```

```{r}
one_team <- df_games %>% 
  filter(team_id == 621) %>% 
  arrange(date) %>% 
  mutate(team_manager_last_game = lag(team_manager)) %>% 
  mutate(manager_change = team_manager_last_game != team_manager)

one_team
# do this for every team
```

```{r}
one_team <- one_team %>% 
  group_by(team_id, temp = rev(cumsum(rev(manager_change)))) %>%
  mutate(games_since_last_change = row_number()) %>%
  ungroup() %>% 
  mutate(games_since_last_change = case_when(manager_change ~ 0, .default = games_since_last_change)) %>% 
  arrange(desc(date)) %>% 
  group_by(team_id, temp = rev(cumsum(rev(games_since_last_change == 0)))) %>%
  mutate(games_to_next_change = row_number()) %>% 
  ungroup() %>% 
  arrange(date)
  
one_team
```

```{r}
one_team <- one_team %>% 
  group_by(team_id, temp = rev(cumsum(rev(manager_change)))) %>%
  mutate(games_since_last_change = row_number()) %>%
  ungroup() %>% 
  mutate(games_since_last_change = case_when(manager_change ~ 0, .default = games_since_last_change)) %>% 
  arrange(desc(date)) %>% 
  group_by(team_id, temp = rev(cumsum(rev(games_since_last_change == 0)))) %>%
  mutate(games_to_next_change = row_number()) %>% 
  ungroup() %>% 
  arrange(date)
  
one_team
```

```{r}
one_team %>% 
  filter(manager_change) %>% 
  select(c(team_name, team_manager, team_manager_last_game, date))
```


## Plot
```{r, echo = FALSE}
one_team %>% 
  filter(season %in% c(2012:2019)) %>% 
  ggplot(mapping = aes(x = date, y = goal_margin, group = team_id)) +
    geom_step() +
    geom_point(data = . %>% filter(manager_change), mapping = aes(colour = manager_change)) +
    geom_label(data = . %>% filter(manager_change), mapping = aes(label = team_manager, y = -6), hjust = 0, check_overlap = T) +
    labs(title = "Goal Margin of Athletic Bilbao over time", y = "Scored - received goals") + 
    guides(color=guide_legend(title="Manager Change"))
```
Add changing manager columns

```{r}
df_games <- df_games %>% 
  arrange(date) %>% 
  group_by(team_id) %>% 
  mutate(team_manager_last_game = lag(team_manager)) %>% 
  mutate(manager_change = team_manager_last_game != team_manager) %>% 
  ungroup()

df_games %>% 
  filter(team_manager != team_manager_last_game)
```

Add counters to isolate before/after treatments

```{r}

df_games_m <- df_games %>% 
  group_by(team_id) %>%
  mutate(games_since_last_change = with(rle(manager_change), sequence(lengths)) * !manager_change) %>%
  #mutate(games_to_next_change = with(inverse.rle(rle(manager_change)), sequence(lengths)) * !manager_change) %>%
  ungroup() # %>% 
  #mutate(games_since_last_change = case_when(manager_change ~ 0, .default = games_since_last_change)) # %>% 
  #arrange(desc(date)) %>% 
  #group_by(team_id, temp = rev(cumsum(rev(games_since_last_change == 0)))) %>%
  #mutate(games_to_next_change = row_number()) %>% 
  #ungroup() %>% 
  #arrange(date)

df_games_m <- df_games_m %>% 
  arrange(desc(date)) %>% 
  group_by(team_id) %>% 
  mutate(games_to_change = with(rle(manager_change), sequence(lengths)) * !manager_change) %>%
  ungroup() %>% 
  arrange(date)

# all changes to check
df_games_m %>%
  filter(team_id == 29) %>%
  select(team_manager, team_manager_last_game, manager_change, games_since_last_change, games_to_change)
```

```{r}
# df_games_m %>%  
#   group_by(team_id) %>% 
#   filter(manager_change)
#   %>% 
#   ggplot(mapping = aes(x = games_since_last_change)) +
#     geom_histogram(binwidth = 10) +
#     labs(title = "How long do managers stay at the same team?", subtitle = "Taking into account all games in this dataset.", x = "# games", y = "contracts")
```

```{r}
TREATMENT_MARGIN <- 3

df_around_treatment <- df_games_m %>% 
  arrange(team_id, date) %>% 
  filter(games_since_last_change <= TREATMENT_MARGIN | games_to_change <= TREATMENT_MARGIN) %>% 
  select(date, team_manager, team_manager_last_game, manager_change, games_since_last_change, games_to_change) %>% 
  ungroup()

df_around_treatment
  
```

make sure we look only at complete treatments (meaning enough matches before/after to compare)

```{r}
df_around_treatment %>% 
  filter(games_to_change <= TREATMENT_MARGIN) %>% 
  mutate(treatment_group = cumsum(as.numeric(manager_change))) %>% 
  select(date, team_manager, team_manager_last_game, manager_change, games_since_last_change, games_to_change, treatment_group) %>% 
  group_by(treatment_group) %>% 
  mutate(treatment_grouper = paste(last(team_manager_last_game), last(team_manager), sep = " - ")) %>% 
  ungroup()
```

```{r}
df_around_treatment %>% 
  filter(games_since_last_change <= TREATMENT_MARGIN) %>% 
  mutate(treatment_group = cumsum(as.numeric(manager_change))) %>% 
  select(date, team_manager, team_manager_last_game, manager_change, games_since_last_change, games_to_change, treatment_group) %>% 
  group_by(treatment_group) %>% 
  mutate(treatment_group_s  = paste(first(team_manager_last_game), first(team_manager), sep = " - ")) %>% 
  ungroup()
```

# completely other approach

```{r}
spanish_league <- df_games %>% 
  filter(competition_id == "ES1") %>% 
  mutate(round_n = as.numeric(sub(" .*", "", round))) %>% 
  group_by(team_id) %>%
  mutate(games_since_last_change = with(rle(manager_change), sequence(lengths)) * !manager_change) %>%
  ungroup() %>% 
  arrange(desc(date)) %>% 
  group_by(team_id) %>% 
  mutate(games_to_change = with(rle(manager_change), sequence(lengths)) * !manager_change) %>%
  ungroup() %>% 
  arrange(date)


spanish_league %>% 
  arrange(team_id, date) %>% 
  #select(date, team_id, team_manager, team_manager_last_game, games_since_last_change, games_to_change, round_n) %>% 
  ggplot(mapping = aes(y = factor(team_name), x = date)) +
    geom_point(shape = "|") +
    geom_point(data = . %>% filter(manager_change), shape = 18, colour = "red", size = 3) +
    labs(title = "Games & manager changes in La Liga per team", subtitle = "Seasons 2011 - 2022")
  
# ---> since/last only for league
```
```{r}

spanish_league_ordered <- spanish_league %>% 
  arrange(team_id, date)
  
spanish_league_ordered %>%
  group_by(team_id) %>% 
  mutate(treatment = paste(team_manager, "->", team_manager_last_game)) %>% 
  mutate(before_treatment = lag(manager_change, 5L)) %>%
  mutate(treatment = lag(treatment, 5L)) %>% 
  ungroup() %>% 
  filter(manager_change | before_treatment) %>% 
  select(date, team_id, team_manager, team_manager_last_game, before_treatment, manager_change, round_n, treatment)
  
```


```{r}

TREATMENT_MARGIN = 3

# get treatment indices
spanish_league_treatment_idx <- which(spanish_league_ordered$manager_change)

# which treatment has a corresponding before treatment?
cond1 <- which(spanish_league_ordered[spanish_league_treatment_idx-TREATMENT_MARGIN,]$games_to_change == TREATMENT_MARGIN)

# which treatment has a corresponding after treatment?
cond2 <- which(spanish_league_ordered[spanish_league_treatment_idx+TREATMENT_MARGIN,]$games_since_last_change == TREATMENT_MARGIN)

# during the season
cond3 <- which(spanish_league_ordered[spanish_league_treatment_idx,]$round_n >= TREATMENT_MARGIN)
cond4 <- which(spanish_league_ordered[spanish_league_treatment_idx,]$round_n <= 38 - TREATMENT_MARGIN)

# get the intersection of ids fullfilling all conditions
spanish_league_treatment_idx_subset <- spanish_league_treatment_idx[intersect(cond1, intersect(cond2, intersect(cond3, cond4)))]

true_real_treatments <- spanish_league_ordered[spanish_league_treatment_idx_subset,] %>% 
  mutate(t = 0)
before_treatment <- spanish_league_ordered[spanish_league_treatment_idx_subset-TREATMENT_MARGIN,] %>% 
  mutate(t = -TREATMENT_MARGIN)
just_before_treatment <- spanish_league_ordered[spanish_league_treatment_idx_subset-1,] %>% 
  mutate(t = -1)
after_treatment <- spanish_league_ordered[spanish_league_treatment_idx_subset+TREATMENT_MARGIN,] %>% 
  mutate(t = TREATMENT_MARGIN)

treatments_df <- rbind(true_real_treatments, before_treatment, after_treatment, just_before_treatment) %>% 
  arrange(team_id, date)

treatments_df %>% 
  group_by(team_id) %>% 
  summarise(count = n()) %>% 
  mutate(validity = count %% 4 == 0)

# treatments_df %>%
#   ggplot(mapping = aes(x = date, y = team_position, colour = factor(t))) +
#     geom_point() +
#     scale_y_reverse()

```
```{r}
treatments_df <- treatments_df %>% 
  mutate(treatment_id = (row_number()-1) %/% 4)
```


```{r, warning=FALSE, message=FALSE}

df_games_m %>% 
  filter(competition_id == "ES1") %>% 
  mutate(round_n = as.numeric(sub(" .*", "", round)),
         team_position = ifelse(games_since_last_change <= 5 | games_to_change <= 5, team_position, NA)) %>% 
  filter(round_n > 5) %>% 
  arrange(team_id, date) %>% 
  ggplot(mapping = aes(x = date, y = team_position, group = team_id, colour = team_name, fill = factor(manager_change))) +
    geom_point(alpha = 0.5, shape = 23) +
    geom_line(alpha = 0.5) +
    scale_y_reverse() +
    guides(colour = FALSE, shape = FALSE, fill = FALSE) +
    geom_text(data = . %>% filter(date == max(date)), mapping = aes(label = team_name), hjust = 0, nudge_x = 5) +
    scale_x_date(expand = expansion(mult = c(0.01, 0.2))) +
    scale_fill_manual(values = c("white", "black", "transparent")) +
    facet_wrap(~season, scales = "free_x")
  
```

```{r}
treatments_df %>% 
  ggplot(mapping = aes(x = date, y = team_position, group = factor(treatment_id), colour = team_name)) +
    geom_point() +
    scale_y_reverse() +
    geom_line() +
    facet_wrap(~treatment_id, scales = "free_x")
```

```{r}

treatments_df <- treatments_df %>% 
  group_by(treatment_id) %>%
  mutate(position_delta = lag(team_position) - team_position) %>% 
  ungroup() %>% 
  mutate(position_delta = ifelse(t == -3, 0, position_delta))

treatments_df
```




```{r}
treatments_df %>% 
  ggplot(mapping = aes(x = t, y = position_delta, group = factor(treatment_id), colour = team_name)) +
    geom_point() +
    geom_line() +
    labs(title = "Change of position, pre and post manager change")
```



```{r}
treatments_df %>% 
  filter(t %in% c(-3, -1)) %>% 
  ggplot(mapping = aes(x = t, y = position_delta)) +
    geom_point() +
    labs(title = "Change of position, pre manager change", subtitle = "On average -1 position change.") +
    geom_smooth(method = lm)
```

### Look for similiar sequences, but without change (treatment)

```{r}
control_df <- spanish_league_ordered %>%
  group_by(team_id) %>% 
  mutate(position_before = lag(team_position, 3L, order_by = team_id,)) %>% 
  mutate(position_delta = position_before - team_position)


control_df %>% 
  filter(position_delta == -1)
```

### There is something wrong below, see validity functions

```{r}

TREATMENT_MARGIN = 3

# get control indices
control_idx <- which(control_df$position_delta == -1)

# during the season
cond1 <- which(control_df[control_idx,]$round_n >= TREATMENT_MARGIN)
cond2 <- which(control_df[control_idx,]$round_n <= 38 - TREATMENT_MARGIN)

# get the intersection of ids fullfilling all conditions
spanish_league_control_idx_subset <- control_idx[intersect(cond1, cond2)]

true_real_controls <- control_df[spanish_league_control_idx_subset,] %>% 
  mutate(t = 0)
before_control <- control_df[spanish_league_control_idx_subset-TREATMENT_MARGIN,] %>% 
  mutate(t = -TREATMENT_MARGIN)
just_before_control <- control_df[spanish_league_control_idx_subset-1,] %>% 
  mutate(t = -1)
after_control <- control_df[spanish_league_control_idx_subset+TREATMENT_MARGIN,] %>% 
  mutate(t = TREATMENT_MARGIN)

controls_df <- rbind(true_real_controls, before_control, just_before_control, after_control) %>% 
  arrange(team_id, date)

controls_df %>% 
  group_by(team_id) %>% 
  summarise(count = n()) %>% 
  mutate(validity = count %% 4 == 0)
  
```

```{r}
controls_df <- controls_df %>% 
  mutate(treatment_id = (row_number()-1) %/% 4)

controls_df <- controls_df %>% 
  group_by(treatment_id) %>%
  mutate(position_delta = lag(team_position) - team_position) %>% 
  ungroup() %>% 
  mutate(position_delta = ifelse(t == -3, 0, position_delta))
```


```{r}
controls_df %>%   
  ggplot(mapping = aes(x = t, y = position_delta, group = factor(treatment_id), colour = team_name)) +
    geom_point() +
    geom_line() +
    labs(title = "Change of position, pre and post manager change")
```

